---
title: "Survival Analysis part 2"
author: "Adam J Sullivan, PhD"
date: "03/21/2018"
output:
   ioslides_presentation: 
    widescreen: true
notes: ''
link: yes
slides: yes
layout: page
css:  "C:/Users/adam_/Dropbox (Personal)/Brown/Teaching/Brown Courses/PHP2511/Spring 2018/website/php-1511-2511.github.io/Notes/slides.css"
---

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(error = TRUE)
opts_chunk$set(warning=FALSE)
opts_chunk$set(message=FALSE)
opts_chunk$set(results="hold")
opts_chunk$set(cache=T)
opts_chunk$set(  tidy=TRUE,size="small")
opts_chunk$set(tidy.opts=list(width.cutoff=60))
options(digits = 3, scipen = 3)
```




# Cox-PH Regression


## Physicians Health Study and Aspirin

Recall the Colorectal Cancer component of the Physicians Health Study

-----------------------------------------------------------
Name        Description
----------- -------------------------------------------
age         Age in years at time of Randomization

asa         0 - placebo, 1 - aspirin

bmi         Body Mass Index (kg/$m^2$)

hypert      1 - Hypertensive at baseline, 0 - Not

alcohol     0 - less than monthly, 1 - monthly to less than daily, 2 - daily consumption
-----------------------------------------------------------

---

-----------------------------------------------------------
Name        Description
----------- -------------------------------------------
dm          0 = No diabetes Mellitus, 1 - diabetes Mellitus

sbp         Systolic BP (mmHg)

exer        0 - No regular, 1 - Sweat at least once per week

csmoke      0 - Not currently, 1 - < 1 pack per day, 2 - $\ge$ 1 pack per day

psmoke      0 - never smoked, 1 - former < 1 pack per day, 2 - former $\ge$ 1 pack per day

pkyrs       Total lifetime packs of cigarettes smoked

crc         0 - No colorectal Cancer, 1 - Colorectal cancer

cayrs       Years to colorectal cancer, or death, or end of follow-up.
------------------------------------------------------------------------

---

For this study each participant contributed 2 pieces of information during follow-up:

1. Information on whether of not they had a Colorectal Cancer(CRC) during follow-up
2. Follow-up time in years, specified as time from randomization until first of
    + end of Study
    + death
    + Colorectal Cancer
    + Loss to follow-up

---    
  

We can load this data into R.
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
phscrc <- read_dta("phscrc.dta")
phscrc <- phscrc %>%
            mutate(age.cat = cut(age, c(40,50,60,70, 90), right=FALSE)) %>%
            mutate(alcohol.use = factor(alcohol>1, labels=c("no", "yes"))) %>%
           mutate(obese = factor(bmi>30, labels=c("Not Obese", "Obese")))
```




## Proportional Hazards Model



The general ***Proportional Hazards Model*** is
$$h(t|X_1,\ldots, X_p) = h_0(t)\exp\left(\beta_1x_1 + \cdots + \beta_px_p\right)$$

 $$\text{or}$$
 
$$\log\left[h(t|X_1,\ldots, X_p)\right] = \log\left[h_0(t)\right] + \beta_1x_1 + \cdots + \beta_px_p$$

where $h_0(t)$ is the baseline hazard function and the "intercept" is $\log\left[h_0(t)\right]$. 


## Semi-Parametric Regression

- Weibull and Exponential are examples of parametric proportional hazards models, where $h_0(t)$ is a specified function. 
- In 1972, Cox generalized these types of models so that we can make inferences on the 
$\beta_1, \ldots,\beta_p$ without specifying $h_0(t)$. 
- We call Cox a semi-parametric regression model
- We fit this using something called *Partial Likelihood Estimation*
- Once again we use an algorithm to maximize the partial likelihood. 



## Interpeting the Model

Let 

- $X=0$ be the control group
- $X=1$ be the treatment group

Then

$$\begin{aligned}
h(t|X=x) &= h_0(t)\exp(\beta x)\\
h(t|X=0) &= h_0(t)\\
&= \text{ baseline hazard for control group}\\
h(t|X=1) &= h_0(t)\exp(\beta)\\
&= \text{ hazard for treated group}\\
\exp(\beta) &= \dfrac{h(t|X=1)}{h(t|X=0)}\\
\end{aligned}$$



## What Does This Mean? 

- This means that the hazard ratio is constant over time (**Proportional Hazards**)
- $\beta$ is the log hazard ratio or log-relative risk
- According to the Cox model

$$\begin{aligned}
\log\left[h]h(t|X=0)\right] &= \log\left[h_0(t)\right]\\
\log\left[h]h(t|X=1)\right] &= \log\left[h_0(t)\right] + \beta\\
\end{aligned}$$

- This means the log of the hazard functions are parallel over time. 
- We make no assumptions about $h_0(t)$. 


##  Verifying Proportional Hazards Assumption

Recall 
$$S(t) = \exp\left(-\Lambda(t)\right)$$
with a binary $X$ we have that

$$\begin{aligned}
\Lambda_1(t) &= \Lambda_0(t)\exp(\beta)\\
S_0(t) &= \exp(\Lambda_0(t))\\
-\log(S_0(t)) &= \Lambda_0(t)\\
\log(-\log(S_0(t))) &=\log(\Lambda_0(t))\\
\end{aligned}$$
$$\begin{aligned}
S_1(t) = exp(-\Lambda_1(t)) &= \exp\left[\Lambda_0(t)\exp(\beta)\right]\\
-\log(S_1(t)) &= \Lambda_0(t)\exp(\beta)\\
\log(-\log(S_1(t))) &= \log(\Lambda_0(t)) + \beta\\
\end{aligned}$$


##  Verifying Proportional Hazards Assumption

- Thus we can see that under the assumption of ***proportional hazards***
    - $\log(-\log(K-M))$ should be parallel over time. 
    - We typically verify this graphically. 
    - Recall the CRC study: 
    

## Example:  Kaplan-Meier Survival

```{r, eval=F}
library(survival)

model <- survfit(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
model

```




## Example:  Kaplan-Meier Survival

```{r, eval=F}
library(survival)

model <- survfit(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
model

```



## Plotting the Kaplan-Meier

```{r, eval=FALSE, message=FALSE, warning=FALSE}

library(survminer)
ggsurvplot(model, legend.labs = 
             c("Non-Drinker", "Drinker"),
           break.time.by=2, fun="cloglog")
```



## Plotting the Kaplan-Meier

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(survminer)
ggsurvplot(model, legend.labs = 
             c("Non-Drinker", "Drinker"),
           break.time.by=2, fun="cloglog")
```


## Cox PH in R

- We can run the Cox PH in R:

```{r, eval=FALSE}
cox.crc <- coxph(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
summary(cox.crc)
```




## Cox PH in R



```{r, echo=FALSE}
cox.crc <- coxph(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
summary(cox.crc)
```



##Interpretation


- This would suggest that the hazard of Colorectal Cancer for those who drink daily is 51% higher than those who drink less than daily.  



## Continuous Example of Cox PH


- Let's consider `age` and `smoking`:

```{R, eval=FALSE}
library(survival)
crc.cox <- coxph(Surv(cayrs, crc) ~ csmok + age , data= subset(phscrc, cayrs>0))
summary(crc.cox)
```




## Continuous Example of Cox PH



```{R, echo=FALSE}
library(survival)
crc.cox <- coxph(Surv(cayrs, crc) ~ csmok + age , data= subset(phscrc, cayrs>0))
summary(crc.cox)
```


## Interpretation 

- Then we could say that for two people with the same smoking status a one year increase in age would lead to an 8.2% increase in the hazard of colorectal cancer with a 95% CI of 6.9% to 9.6%. 
- We would also be able to say that for 2 people the same age, a person who is a current smoker would have a 37% increase in hazard of colorectal cancer than a non smoker. 





# Cox-PH Regression


## Assessing Diagnostics and Model Fit with Cox-PH

- With the Cox PH model we will consider 2 things
    - Checking Proportional Hazards Assumption
    - Checking for Influential Observations
    


## The Data

- We will consider Recidivism of 432 male patients. 
- They all were observed for 1 year prior to release from prison.
- The following slides will contain the variables. 

## Variables



----------------------------------------------------
Variable      Description
------------  ----------------------------------------
week          week of first arrest after release, or censoring time.

arrest        the event indicator, equal to 1 for those arrested during the period of the study and 0 for those who were not arrested.

fin           a factor, with levels yes if the individual received financial aid after release from prison, and no if he did not; financial aid was a randomly assigned factor manipulated by the researchers.
------------------------------------------------------


## Variables


----------------------------------------------------
Variable          Description
----------------  ----------------------------------------
age               in years at the time of release.

wexp              a factor with levels yes if the individual had full-time work experience prior to incarceration and no if he did not.

mar               a factor with levels married if the individual was married at the time of release and not married if he was not.

paro              a factor coded yes if the individual was released on parole and no if he was not. 

prior             number of prior convictions.
-------------------------------------------------------




## Variables


----------------------------------------------------
Variable          Description
----------------  ----------------------------------------
educ              education, a categorical variable coded numerically, with codes 2 (grade 6 or less), 3 (grades 6 through 9), 4 (grades 10 and 11), 5 (grade 12), or 6 (some post-secondary).6

emp1 – emp52      factors coded yes if the individual was employed in the corresponding week of the study and no otherwise.


race              a factor with levels black and other.
----------------------------------------


## Reading Data in

```{r}
url <- "http://socserv.mcmaster.ca/jfox/Books/Companion/data/Rossi.txt"
Rossi <- read.table(url, header=TRUE)
```




## Our Model

```{r, eval=FALSE}
library(survival)
mod1 <- coxph(Surv(week,arrest)~ fin + age + race + 
                wexp + mar + paro + prio, 
              data=Rossi)
tidy1 <- tidy(mod1, exponentiate=T)
knitr::kable(tidy1[-c(3,4)])
```





## Our Model

```{r, echo=FALSE}
library(survival)
mod1 <- coxph(Surv(week,arrest)~ fin + age + race + 
                wexp + mar + paro + prio, 
              data=Rossi)
tidy1 <- tidy(mod1, exponentiate=T)
knitr::kable(tidy1[-c(3,4)])
```



## Plotting Regression


```{r, eval=F}
library(ggplot2)
library(ggfortify)
autoplot(survfit(mod1), surv.linetype = 'dashed', surv.colour = 'blue',
         conf.int.fill = 'dodgerblue3', conf.int.alpha = 0.5, censor = FALSE)
```




## Plotting Regression


```{r, echo=F}
library(ggplot2)
library(ggfortify)
autoplot(survfit(mod1), surv.linetype = 'dashed', surv.colour = 'blue',
         conf.int.fill = 'dodgerblue3', conf.int.alpha = 0.5, censor = FALSE)
```




## Checking Proportional Hazards

- We have tested these before with Schoenfeld Residuals
- We can do this with the `cox.zph()` function. 

## Checking Proportional Hazards

```{R}
cox.zph(mod1)
```


## Conclusion

- We see there is an issue
    - `age` is an issue
    - `wexp` is an issue as well. 
- What do we do???


## Enter Stratification


- We can adjust for a variable that does not meet the proportional hazards assumption by stratification. 
- Assume we have $Z$ which does not allow for PH
$$h(t|X, Z=j) = h_j(t)exp(X\beta)$$
- $j=1,ldots,C$ levels of Z. 


## Create Age Categories


```{R}
Rossi$age.cat <- cut(Rossi$age, c(0, 19, 25, 30, Inf))
xtabs(~age.cat, data=Rossi)
```

## Re Run the Model

```{r}
mod2  <- coxph(Surv(week,arrest)~ fin + race + 
                mar + paro + prio + strata(wexp, age.cat), 
              data=Rossi)
tidy2 <- tidy(mod2, exponentiate=T)
knitr::kable(tidy2[-c(3,4)])
```




## PH Assumption

```{r}
cox.zph(mod2)
```




## Influential Observations


- to test this let's build a smaller model

```{r}
mod3 <-  coxph(Surv(week, arrest) ~ fin + 
                 prio + strata(age.cat), data=Rossi)
```



## Influential Observations

- With Cox PH we will use DFBETA to tell.
- DFBETA's measure how much an observation has effected the estimated coefficient. 
- We look for values to be under $\dfrac{2}{\sqrt{n}}$. 

## DFBETA in R


```{r}
library(survminer)
2/sqrt(dim(Rossi)[1])
```
```{r, eval=F}
ggcoxdiagnostics(mod3,  type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```


## DFBETA in R

```{r, echo=F}
ggcoxdiagnostics(mod3,  type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```




