---
title: "Survival Analysis"
author: "Adam J Sullivan, PhD"
date: "03/19/2018"
output:
   ioslides_presentation: 
    widescreen: true
notes: ''
link: yes
slides: yes
layout: page
css:  "C:/Users/adam_/Dropbox (Personal)/Brown/Teaching/Brown Courses/PHP2511/Spring 2018/website/php-1511-2511.github.io/Notes/slides.css"
---

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(error = TRUE)
opts_chunk$set(warning=FALSE)
opts_chunk$set(message=FALSE)
opts_chunk$set(results="hold")
opts_chunk$set(cache=T)
opts_chunk$set(  tidy=TRUE,size="small")
opts_chunk$set(tidy.opts=list(width.cutoff=60))
options(digits = 3, scipen = 3)
```




# Binary Covariate - Survival Time Outcome

## Binary Covariate - Survival Time Outcome


- For this problem we are looking at a survival time outcome. 
- This type of outcome has both event status as well as a time included. 
- For this type we might be comparing how long after taking a medication that a patient is cured.
- We will look at various ways to do this now. 




## Goals of Survival Analysis

- Goals of Survival Analysis:
    + Estimate distribution of of survival time for a population
    + Test for equality of survival distributions among 2 or more groups
        - Control
        - Treated
    + Estimate the absolute or relative treatment effects
    + Estimate and control for effects of other covariates
        - Confounding
        - Effect Modification/Interaction
    + Find confidence intervals and significance for effects. 





## Describing and Characterizing Survival Data

- The event
    + What is the event of interest? 
    + How is it specifically defined?
- The Origin
    + What is the initial starting point?
    + This must be before anyone in the study has had the event of interest. 
- The Metric for time
    + What is the scale in which events are recorded. 



  
## Examples
- Time to relapse after end of treatment among cancer patients.
    + *Event*: Relapse of Cancer
    + *Origin*: End of Treatment.
    + *Time Metric*: Days
    
    
- Length of stay in hospital for patients who suffered a heart attack.
    + *Event*: Leaving Hospital.
    + *Origin*: Admission to Hospital.
    + *Time Metric*: Hours.

## Examples


- Age of onset of breast cancer in individuals with family history.
    + *Event*: Onset of Breast Cancer.
    + *Origin*: Birth.
    + *Time Metric*: Years




## The Survival Function

- The random variable of interest, $T$ is the time to the event of interest.
- We then know that $T$ is positive and by definition:
$$T\ge0$$
The Survival function, $S(t)$, is the proportion of individuals who have not experienced at event at some time $t>0$.This is defined by:
$$S(t) = \Pr(T>t)$$




## The Survival Function



- If the event of interest is death, this would mean the subject is still alive at time $t$. 
- Then $S(t)$ would be the proportion of subjects alive at time $t$. 
- This simple proportion is for when there is no *censoring* (discussed later on).



##Features of $S(t)$

- A survivor function is a sequence of probabilities up till time $t$
$$0\le S(t)\le1, \qquad \text{ for each } t\ge0$$

- At time, $t=0$, $S(0)=1$.
- $S(t)$ decreases as events happen over $t$
    + So that if $t_2\ge t_1$ then $S(t_2)\le S(t_1)$. 
    + They are non-increasing functions.
    
##Features of $S(t)$


- For large, $t$, such as $t=\infty$, $S(t)$ goes to 0. 
    + This means that for some events $S(t)$ approaches a 0 asymptote.
    + However for some diseases, some people may be cured so that $S(t)$ approaches a non-zero asymptote. 
- Graphical displays are a common method to display summaries survivor functions. 





## The Hazard Function 

The Hazard function is defined as the instantaneous rate of failure at time $t$, given that a subject as survived up until time $t$. 

It is also referred to as:

- Hazard Rate.
- Failure Rate.
- Mortality Rate

##Relationship between $h(t)$ and $S(t)$:

- The hazard function $(h(t))$, survival function $(S(t))$, probability density function $(f(t))$, and cumulative distribution function $(F(t))$ are all related.
- They are defined in terms of random variable $T$ which is the time until event. 
- In censored subjects we only know that $T>t$ for a subject censored at time $t$. 
- Mathematically:
$$h(t) = \dfrac{f(t)}{S(t)} = -\dfrac{\dfrac{d}{dt}S(t))}{S(t)} = -\dfrac{d}{dt}\log\left[S(t)\right]$$




##Kaplan-Meier Estimator
The control group was easy to analyze as it had no censoring so we could calculate it by hand. However with the introduction to censoring we need a new estimator. 

- Kaplan-Meier is a non-parametric method.
    + No assumptions of distribution
- We define patients to be at risk at time $t$ if they have not experienced the event just before time $t$ and are not yet censored just before time $t$. 




## Log Rank Test

- The Logrank Test is a hypothesis test for 2 or more independent samples of survival data. 
- The hypothesis being tested are:

$$H_0: S_1(t) = S_2(t) \text{ for all }t$$
$$\text{and}$$
$$H_1: S_1(t) \ne  S_2(t) \text{ for some }t$$


## Log Rank Test


If $H_0$ is true then

- $h_1(t)=h_2(t) \text{ for all }t$
- $\Lambda_1(t)=\Lambda_2(t)  \text{ for all }t$



## How do we calculate this test statistic?

1. Construct a 2x2 table at the time of each observed failure. 
2. Calculate the Mantel-Haenszel chi-square test statistic. 

- We have $K$ distinct observed failure times:
$$t_1<\cdots<t_K$$
- at the $i^{\text{th}}$ observed failure time $t_i$:



## How do we calculate this test statistic?

Treatment | Died | Alive | At Risk
--------- | ---- | ----- | ---------
Control | $a_i$ | $b_i$ | $n_{1i}$
Treated | $c_i$ | $d_i$ | $n_{2i}$
total | $m_{1i}$ | $m_{2i}$ | $n_i$


where
$$\begin{aligned}
n_{1i} &= \text{ number at risk at } t_i \text{ from Control}\\
n_{2i} &= \text{ number at risk at } t_i \text{ from Treated}\\
m_{1i} &= \text{ number of failures at } t_i\\
m_{2i} &= \text{ number surviving past } t_i\\
n_i &= \text{ total numer at risk at } t_i\\
\end{aligned}$$


## Relation to Mantel-Haenszel Test

- This test is exactly the same as a Mantel-Haenszel test applied to $K$ strata
$$\chi^2_{MH} = \dfrac{\left[\sum_{i=1}^K (a_i - E(a_i))\right]^2}{\sum_{i=1}^K Var(a_i)}$$
- where
$$\begin{aligned}
E(a_i) &= \dfrac{n_{1i}m_{1i}}{n_i}\\
Var(a_i) &= \dfrac{n_{1i}n_{2i}m_{1i}m_{2i}}{n_i^2(n_i-1)}\\
\end{aligned}$$



## Relation to Mantel-Haenszel Test

We compute the expectation that the null hypothesis is true and there is no difference in survival between the groups. We consider all margins fixed but $a_i$ is random and thus we have a hypergeometric distribution. 

- Under $H_0$ we have that $S_1(t)=S_2(t)$ and this means
    + $\chi^2_{MH}\sim \chi^2_1$
    + Reject $H_0$ when $\chi^2_{MH}>\chi^2_{1,1-\alpha}$
- This test is most powerful if the hazard ratio is constant over time. 
- We can easily extend this to compare 3 or more independent groups.




## Recall the Colorectal Cancer component of the Physicians Health Study

-----------------------------------------------------------
Name        Description
----------- -------------------------------------------
age         Age in years at time of Randomization

asa         0 - placebo, 1 - aspirin

bmi         Body Mass Index (kg/$m^2$)

hypert      1 - Hypertensive at baseline, 0 - Not

alcohol     0 - less than monthly, 1 - monthly to less than daily, 2 - daily consumption
-----------------------------------------------------------

##

-----------------------------------------------------------
Name        Description
----------- -------------------------------------------
dm          0 = No diabetes Mellitus, 1 - diabetes Mellitus

sbp         Systolic BP (mmHg)

exer        0 - No regular, 1 - Sweat at least once per week

csmoke      0 - Not currently, 1 - < 1 pack per day, 2 - $\ge$ 1 pack per day

psmoke      0 - never smoked, 1 - former < 1 pack per day, 2 - former $\ge$ 1 pack per day

pkyrs       Total lifetime packs of cigarettes smoked

crc         0 - No colorectal Cancer, 1 - Colorectal cancer

cayrs       Years to colorectal cancer, or death, or end of follow-up.
------------------------------------------------------------------------

## What Each Subject Contributed



1. Information on whether of not they had a Colorectal Cancer(CRC) during follow-up
2. Follow-up time in years, specified as time from randomization until first of
    + end of Study
    + death
    + Colorectal Cancer
    + Loss to follow-up

## Loading Data

```{r}
library(tidyverse)
library(haven)
phscrc <- read_dta("phscrc.dta")
phscrc <- phscrc %>%
            mutate(age.cat = cut(age, c(40,50,60,70, 90), right=FALSE)) %>%
            mutate(alcohol.use = factor(alcohol>1, labels=c("no", "yes"))) %>%
           mutate(obese = factor(bmi>30, labels=c("Not Obese", "Obese")))


```



## Example:  Kaplan-Meier Survival


```{r, eval=F}
library(survival)

model <- survfit(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
model

```



## Example:  Kaplan-Meier Survival


```{r, echo=F}
library(survival)

model <- survfit(Surv(cayrs,crc) ~alcohol.use,
          data = subset(phscrc, cayrs>0))
model
```


## Plotting the Kaplan-Meier


```{r, eval=F}
library(survminer)
ggsurvplot(model, conf.int=FALSE, risk.table=TRUE,
           risk.table.col="strata", legend.labs = 
             c("Non-Drinker", "Drinker"), break.time.by=2, ylim=c(0.95,1))
```


## Plotting the Kaplan-Meier


```{r, echo=F}
library(survminer)
ggsurvplot(model, conf.int=FALSE, risk.table=TRUE,
           risk.table.col="strata", legend.labs = 
             c("Non-Drinker", "Drinker"), break.time.by=2, ylim=c(0.95,1))
```




## Log Rank Test


```{r, eval=F}
model <- survdiff(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
model
```



## Log Rank Test


```{r, echo=F}
model <- survdiff(Surv(cayrs,crc) ~ alcohol.use,
          data = subset(phscrc, cayrs>0))
model
```

## Conclusion

- What can we conclude?





# Categorical Covariate - Survival Time Outcome



## Categorical Covariate - Survival Time Outcome


- Same exact method Kaplan-Meier and Log Rank.
- We will see an example of this.



## CRC Example:  Kaplan-Meier Survival


```{r, eval=F}

model <- survfit(Surv(cayrs,crc) ~csmok,
          data = subset(phscrc, cayrs>0))
model

```




## PBC-3 Example:  Kaplan-Meier Survival


```{r, echo=F}
model <- survfit(Surv(cayrs,crc) ~csmok,
          data = subset(phscrc, cayrs>0))
model
```


## PBC-3 Example: Plotting the Kaplan-Meier


```{r, eval=F}
ggsurvplot(model, conf.int=TRUE, risk.table=TRUE,
           risk.table.col="strata", break.time.by=2, ylim=c(0.95,1))
```




## Log Rank Test


```{r, eval=F}
model <- survdiff(Surv(cayrs,crc) ~ csmok,
          data = subset(phscrc, cayrs>0))
model
```

## Log Rank Test


```{r, echo=F}
model <- survdiff(Surv(cayrs,crc) ~ csmok,
          data = subset(phscrc, cayrs>0))
model
```

## Conclusion
