---
title: "Lab 3 - Linear Regression Final Model"
author: "Your Name"
output: 
  html_document
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=F)
knitr::opts_chunk$set(eval=T)
```


## Introduction


We will consider data from Life In America's Small Cities, By G.S. Thomas. Our goal will be to model the total crime. 

```{r}
library(readxl)
crimes <- read_excel("crimes.xlsx")

```

### Variables


------------------------------------------
Name          Description
------------  ------------------------------
X1            total overall reported crime rate per 1 million residents

X2            reported violent crime rate per 100,000 residents

X3            annual police funding in $/resident

X4            % of people 25 years+ with 4 yrs. of high school

X5            % of 16 to 19 year-olds not in highschool and not highschool graduates.

X6            % of 18 to 24 year-olds in college

X7            % of people 25 years+ with at least 4 years of college
---------------------------------------------------------------

### Data Cleaning

Let's use some better names:

```{r}
library(tidyverse)
crimes <- crimes %>%
  rename( tot_crime = X1,
          rep_crime = X2,
          pol_fund = X3, 
          `25plus_hs` = X4,
          not_hs_grad = X5,
          in_college = X6,
          college = X7  )
```

##Data Exploration

We should always begin with a data exploration. 

1. Augment the shiny code from lab 2 in order to explore this data. 

```{r, eval=TRUE}
library(shiny)
library(ggplot2)
data <- crimes
shinyApp(
  




 
# Define UI for iris application
shinyUI(pageWithSidebar(
 
  # Application title
  headerPanel(""),
 
  # Sidebar with controls to select the variable to plot against mpg
  # and to specify whether outliers should be included
  sidebarPanel(
    selectInput("variable", "First variable:",
                list(
"tot_crime", "rep_crime", "pol_fund", "25plus_hs", "not_hs_grad", "in_college", "college"
                                          )),
 
    selectInput("variable2", "Second variable:",
                list(
"tot_crime", "rep_crime", "pol_fund", "25plus_hs", "not_hs_grad", "in_college", "college"
                                          ))
  ),


  mainPanel(
     h3(textOutput("caption")),
     plotOutput("plot")
  )
)),


 

 
# Define server logic required to plot variables
shinyServer(function(input, output) {
 
  # Create a reactive text
  text <- reactive({
     paste(input$variable, 'versus', input$variable2)
  })
 formulaText <- reactive({
    paste(input$variable, "~", input$variable2)
  })
  # Return as text the selected variables
  output$caption <- renderText({
    text()
  })
 
  # Generate a plot of the requested variables
 output$plot <- renderPlot({
     p <- ggplot(data, aes_string(x=input$variable, y=input$variable2)) + geom_point()
     print(p)
  })
}),

 options = list(height = 600)
  
)
```

2. Consider a correlation matrix:

```{r}
library(GGally)
ggcorr(crimes, colour="gray20")
```

3. Consider a Scatterplot matrix:

```{r,fig.width=8, fig.height=10}
ggpairs(crimes)
```



4. Consider Simple Linear Regressions of Average Crime
```{r}
library(broom)
library(tidyverse)
library(knitr)
mod1 <- lm(tot_crime~rep_crime, data=crimes)
mod2 <- lm(tot_crime~pol_fund, data=crimes)
mod3 <- lm(tot_crime~`25plus_hs`, data=crimes)
mod4 <- lm(tot_crime~not_hs_grad, data=crimes)
mod5 <- lm(tot_crime~in_college, data=crimes)
mod6 <- lm(tot_crime~college, data=crimes)

tidy1 <- tidy(mod1)[-1, -c(3:4)]
tidy2 <- tidy(mod2)[-1, -c(3:4)]
tidy3 <- tidy(mod3)[-1, -c(3:4)]
tidy4 <- tidy(mod4)[-1, -c(3:4)]
tidy5 <- tidy(mod5)[-1, -c(3:4)]
tidy6 <- tidy(mod6)[-1, -c(3:4)]

kable(bind_rows(tidy1,tidy2,tidy3,tidy4,tidy5,tidy6))

```

