---
title: "Lab 3 - Linear Regression Final Model"
author: "Your Name"
output: 
  html_document
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=F)
knitr::opts_chunk$set(eval=T)
```


## Introduction


We will consider data from Life In America's Small Cities, By G.S. Thomas. Our goal will be to model the total crime. 

```{r}
library(readxl)
crimes <- read_excel("crimes.xlsx")

```

### Variables


------------------------------------------
Name          Description
------------  ------------------------------
X1            total overall reported crime rate per 1 million residents

X2            reported violent crime rate per 100,000 residents

X3            annual police funding in $/resident

X4            % of people 25 years+ with 4 yrs. of high school

X5            % of 16 to 19 year-olds not in highschool and not highschool graduates.

X6            % of 18 to 24 year-olds in college

X7            % of people 25 years+ with at least 4 years of college
---------------------------------------------------------------

### Data Cleaning

Let's use some better names:

```{r}
library(tidyverse)
crimes <- crimes %>%
  rename( tot_crime = X1,
          rep_crime = X2,
          pol_fund = X3, 
          `25plus_hs` = X4,
          not_hs_grad = X5,
          in_college = X6,
          college = X7  )
```

##Data Exploration

We should always begin with a data exploration. 

1. Augment the shiny code from lab 2 in order to explore this data. 

2. Consider a correlation matrix:

```{r}
library(GGally)
ggcorr(crimes, colour="gray20")
```

3. Consider a Scatterplot matrix:

```{r}
ggpairs(crimes)
```



4. Consider Simple Linear Regressions of Average Crime
```{r}
library(broom)
library(tidyverse)
library(knitr)
mod1 <- lm(tot_crime~rep_crime, data=crimes)
mod2 <- lm(tot_crime~pol_fund, data=crimes)
mod3 <- lm(tot_crime~`25plus_hs`, data=crimes)
mod4 <- lm(tot_crime~not_hs_grad, data=crimes)
mod5 <- lm(tot_crime~in_college, data=crimes)
mod6 <- lm(tot_crime~college, data=crimes)

tidy1 <- tidy(mod1)[-1, -c(3:4)]
tidy2 <- tidy(mod2)[-1, -c(3:4)]
tidy3 <- tidy(mod3)[-1, -c(3:4)]
tidy4 <- tidy(mod4)[-1, -c(3:4)]
tidy5 <- tidy(mod5)[-1, -c(3:4)]
tidy6 <- tidy(mod6)[-1, -c(3:4)]

kable(bind_rows(tidy1,tidy2,tidy3,tidy4,tidy5,tidy6))

```

5. Let's say we are interested in specifically understanding the effect that expenditure has on SAT score. We have seen above that this seems to have a negative relationship. 

```{r}
library(car)
library(olsrr)
mmps(mod1)
ols_rvsp_plot(mod1)

mod7 <- lm(sat~expend*ratio, sat)
```