```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=F)
```



## The Data 


We will begin with an example of simple logistic regression with a sample of the [Western Collaborative Group Study](https://clinicaltrials.gov/ct2/show/NCT00005174).  This study began in 1960 with 3154 men ages 39-59, who were employed in one of 11 California based companies.  They were followed until 1969 during this time, 257 of these men developed coronary heart disease (CHD).  You can read this data in with the code below. [Lab 4 Markdown](https://php-1511-2511.github.io/labs/lab4.Rmd)

## Reading in the Data

```{r, tidy=TRUE}

library(haven)
wcgs <- read_dta("wcgs2.dta")
wcgs <- wcgs[,-16]
```

## The Variables


-----------------------------------------------------------
Name    Description
------- -------------------------------------------
id      Subject identification number


age     Age in years

height  Height in inches

weight  Weight in lbs.

sbp     Systolic blood pressure in mm 

dbp     Diastolic blood pressure in mm Hg

chol    Fasting serum cholesterol in mm 

behpat  Behavior

        1 = A1

        2 = A2

        3 = B3

        4 = B4

ncigs   Cigarettes per day

dibpat  Behavior

        1 = type A

        2 = type B

chd69   Coronary heart disease

        1 = Yes

        0 = no

typechd Type of CHD

        1 = myocardial infarction or death

        2 = silent myocardial infarction

        3 = angina perctoris

time169 Time of CHD event or end of follow-up

arcus   Arcus senilis

        0 = absent

        1 = present

bmi     Body Mass Index
-----------------------------------------------------------



1. Begin by exploring the data and relationships further. What kind of relationships do you notice with outcome of Coronary Heart Disease (CHD)? Make sure to use appropriate tables, graphs or summaries. 

**For continuous covariates, we can explore their relationship with CHD by using boxplots. For categorical covariates, we can create contingency tables by `table` function, in order to explore their relationship with CHD.**

```{r,echo=T,eval=T,results='hide'}
numeric <- c("age","height","weight","sbp","dbp","chol","ncigs","bmi","time169")
knitr::kable(summary(wcgs[numeric]))
category <- c("behpat","dibpat","chd69","typchd69","arcus")
apply(wcgs[category],2,table)
```

```{r}
#plot for numerical variables vs. CHD
library(ggplot2)
par(mfrow = c(3,3))
for (var in numeric) {
  boxplot(as.formula(paste0(var,"~chd69")), data = wcgs, xlab = "chd", ylab = var)
}
```

```{r}
#tables for categorical variable vs. CHD
for (var in category) {
  print(table(wcgs[c("chd69",var)]))
}
```

2. Create logistic regressions with the following variables:
    - `age`
    - `bmi`
    - `sbp`
    - `dbp`
    - `behpat`
    - `ncigs`
    - `dibpat`
    - `arcus`
    
    Create a table with these univariate regressions. Interpret them. 
    
```{r}
#turn categorical variables into factors
library(tidyverse)
wcgs<-wcgs %>% mutate(behpat=factor(behpat,levels = c(1,2,3,4),labels=c("A1","A2","B3","B4"))) %>%
           mutate(dibpat=factor(dibpat,levels=c(1,0),labels=c("A","B")))%>%
           mutate(chd69=as.factor(chd69))%>%
           mutate(typchd69=as.factor(typchd69))%>%
           mutate(arcus=as.factor(arcus))

```

```{r}
library(broom)
result <- c()
for (var in c("age","bmi","sbp","dbp","behpat","ncigs","dibpat","arcus")) {
  f <- as.formula(paste0("chd69~", var))
  m <- glm(f, data = wcgs, family = binomial(link = "logit"))
  r <- tidy(m, exponentiate = T, conf.int = T)[-1,-c(3,4)]
  result <- rbind(result,r)
}
result
```
**The coefficient interpretation is different for continuous covariate and categorical covariate. For a continuouse covariate, its exponentiated coefficient can be interpreted in following ways:** 

**Take `age` as an example:** 

**$\cdot$ If there are 2 people who differ in age by one year the older person would on average have an odds of CHD 1.077 times that of the younger.**

**$\cdot$ On average a person has an odds of CHD 0.077 higher than someone a year younger.**

**$\cdot$ On average a person has an odds of CHD 7.726% higher than someone a year younger.**


**For a categorical covariate, its exponentiated coefficeint can be interpreted in following ways:**

**Take `arcus` as an example:**

**$\cdot$ On average a person with Arcus Senilis present has an odds of CHD 1.635 times that of the odds of CHD for another person with absence of Arcus Senilis.**

**$\cdot$ On average a person with Arcus Senilis present has an odds of CHD 63.528% higher than the odds of CHD for another person with absence of Arcus Senilis.**


3. Consider a multiple with all of the above variables. 

```{r}
table(wcgs$dibpat,wcgs$behpat)
m1<-glm(chd69 ~ age+bmi+sbp+dbp+behpat+dibpat+ncigs+arcus, data = wcgs, family = binomial(link = "logit"))

#m2<-glm(chd69 ~ age+bmi+sbp+dbp+dibpat+behpat+ncigs+arcus, data = wcgs, family = binomial(link = "logit"))

tidy(m1, exponentiate = T, conf.int = T)[,-c(3,4)]
#tidy(m2, exponentiate = T, conf.int = T)
```
    
    a. Comment on the changes you see in coefficients from the univariate to multiple regression model. 
    
  **From the output of m1, we can see there's no coefficent for `dibpat`. This is due to `behpat` and `dibpat` are correlated, in the sense if behpat is A1 or A2, then dibpat must be A. Therefore, `bahpat` contains the information of `dibpat`. For other covariates, we can see the predictor `sbp` is no longer statistically significant in the mutiple regression model. The coefficeint for `behpatA2` increased from 0.98 to 1.03, the coefficient for `arcus` decreased from 1.63 to 1.37.** 
    
    
    b. Pick the 3 most significant coefficients and interpret them. 

  **Based on p-values, we choose `age`, `ncigs` and `sbp`` as 3 most significant coefficients.** 
  
  **Take `age` as an example: On average a person has an odds of CHD 0.060 higher than someone a year younger, conditioning on all other covariates.**
  

4. What is the log odds of CHD for a 65 year old with bmi of 36, sbp of 136, dbp 78, behpat A2, ncigs=3, dibpat Type B and No Arcus. 

**For prediction, we need to first create a new data entry, then use `predict` function with the new dataset to predict the log odds.**

```{r}
new_input <- data.frame(age = 65,bmi = 36, sbp = 136,dbp = 78,behpat = "A2",ncigs = 3,arcus = factor(0), dibpat = "B")
log_odds <- predict(m1,newdata=new_input)
log_odds
```

5. What is the probability of CHD for a 65 year old with bmi of 36, sbp of 136, dbp 78, behpat A2, ncigs=3, dibpat Type B and No Arcus. 
```{r}
p <- exp(log_odds)/exp(1+log_odds)
p
```
