---
title       : Poisson Regression
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(error = TRUE)
opts_chunk$set(warning=FALSE)
opts_chunk$set(message=FALSE)
opts_chunk$set(results="hold")
opts_chunk$set(cache=F)
opts_chunk$set(  tidy=F,size="small")
opts_chunk$set(tidy.opts=list(width.cutoff=60))
options(digits = 3, scipen = 3)
```




# Poisson Regression


--- .class #id

## Poisson Regression

- Another situation where we can use a GLM is with Poisson data. 
- The poisson model fits data where
    - Response is a count that follows a Poisson distribution. 
    - If the events are recurrent than the probability of a 2$^\text{nd}$ event must not have an increase over the probability of the 1$^\text{st}$ event. (This would fail for say blood clots. )
    - Incidence rates remain constant over time.  
    - Incidence rate multiplied by exposure gives the expected number of events. 
    - Over a very small exposure time $t$ the probability of more than one event happening is 0. 

--- .class #id

## Poisson Regression Strengths

- Poisson regression generalizes crude and stratified incidence rates. 
- Does not require that subjects are followed for the same amount of time
    + This can be a weakness of logistic regression when we assume that subjects are followed for the same amount of time given that most studies never achieve this. 

--- .class #id 

## Link Function

- With Poisson Regression it can be shown that we are interested in 
$$\log\left(\mu\right) =  \beta_0 + \beta_1x_1 + \cdots + \beta_px_p$$
- This would show us we would be using the $\log$ link. 

--- .class #id
 

## Example

- Unlike Logistic Regression where we took more time to go through the math the benefit of using a GLM is that we no longer need to discuss how to estimate our coefficients. 
- We once again will by using maximum likelihood theory. 
- We will instead discuss Poisson Regression through example. 



--- .class #id

## Physician Health Study Variables



| Name | Description | 
| ----- | ------------- | 
| age  |       Age in years at time of Randomization |
| asa   |      0 - placebo, 1 - aspirin | 
| bmi    |     Body Mass Index (kg/$m^2$) | 
| hypert  |    1 - Hypertensive at baseline, 0 - Not | 
| alcohol  |   0 - less than monthly, 1 - monthly to less than daily, 2 - daily consumption |
| dm      |    0 = No diabetes Mellitus, 1 - diabetes Mellitus |
| sbp      |   Systolic BP (mmHg) | 

--- .class #id


## Physician Health Study Variables

| Name | Description | 
| ----- | ------------- | 
| exer    |    0 - No regular, 1 - Sweat at least once per week | 
| csmoke  |    0 - Not currently, 1 - < 1 pack per day, 2 - $\ge$ 1 pack per day | 
| psmoke  |    0 - never smoked, 1 - former < 1 pack per day, 2 - former $\ge$ 1 pack per day |
| pkyrs   |    Total lifetime packs of cigarettes smoked | 
| crc     |    0 - No colorectal Cancer, 1 - Colorectal cancer | 
| cayrs   |    Years to colorectal cancer, or death, or end of follow-up. | 

--- .class #id


## What Each Subject Contributed



1. Information on whether of not they had a Colorectal Cancer(CRC) during follow-up
2. Follow-up time in years, specified as time from randomization until first of
    + end of Study
    + death
    + Colorectal Cancer
    + Loss to follow-up

--- .class #id


## Loading Data

```{r}
library(tidyverse)
library(haven)
phscrc <- read_dta("phscrc.dta")
phscrc <- phscrc %>%
            mutate(age.cat = cut(age, c(40,50,60,70, 90), right=FALSE)) %>%
            mutate(alcohol.use = factor(phscrc$alcohol>0, labels=c("no", "yes")))


```


--- .class #id

## Alcohol Use by Age
    
We then can consider the following table of information
    
<style type="text/css">
	table.tableizer-table {
		font-size: 12px;
		border: 1px solid #CCC; 
		font-family: Arial, Helvetica, sans-serif;
		 width: 80%;
    margin-left:10%; 
    margin-right:10%;
	} 
	.tableizer-table td {
		padding: 4px;
		margin: 3px;
		border: 1px solid #CCC;
	}
	.tableizer-table th {
		background-color: #104E8B; 
		color: #FFF;
		font-weight: bold;
	}
</style>
<table class="tableizer-table">
<thead><tr class="tableizer-firstrow"><th></th><th>Alcohol Users</th><th>Non-Alcohol Users</th></tr></thead><tbody>
 <tr><td>Ages</td><td>$\dfrac{\text{Events(MI)}}{\text{Person-Years}}$</td><td>$\dfrac{\text{Events(MI)}}{\text{Person-Years}}$</td></tr>
 <tr><td>40-49</td><td>$\dfrac{8}{69.723}=0.1147$</td><td>$\dfrac{31}{208.093}=0.1490$</td></tr>
 <tr><td>50-59</td><td>$\dfrac{21}{172.485}=0.1217$</td><td>$\dfrac{59}{426.540}=0.1383$</td></tr>
 <tr><td>60-69</td><td>$\dfrac{32}{233.063}=0.1373$</td><td>$\dfrac{62}{410.415}=0.1511$</td></tr>
 <tr><td>70+</td><td>$\dfrac{20}{121.789}=0.1642$</td><td>$\dfrac{21}{129.177}=0.1626$</td></tr>
 <tr><td>Total </td><td>$\dfrac{81}{597.060}=0.13566$</td><td>$\dfrac{173}{1174.225}=0.1473$</td></tr>
</tbody></table>


--- .class #id


## Reasoning for Poisson Regression

- Before we continue we will discuss why we may use Poisson regression here rather than logistic. 
    - Poisson regression is used to model expected number of events given covariates. 
    - We can use either categorical or continuous covariates. 
    - The number of events for each subject is independent from subject to subject and each subject has a distribution:
$$Y_i \approx Poisson(\mu_i=\lambda_it_i)$$
        + This incidence rate ($\lambda_i$) of CRC is constant over time but may vary individually based on covariates for subject $i$. 


--- .class #id
    
## What do we have?

-  For $i^{\text{th}}$ subject we have a follow up of $t_i$ years $i=1,\ldots,22071$. 
$$
Y_i = \begin{cases}
1 & \text{ if patient develops CRC}\\
0 & \text{ Otherwise}\\
\end{cases}
$$
- $Y_i$ is not binomial since $t_i$ is different for every subject. 
- We can then assume that $Y_i$ is Poisson

$$
\begin{aligned}
E(Y_i) &= \mu_i = \lambda_it_i\\
\log(\mu_i) &= \log(\lambda_i) + \log(t_i)\\
&= \beta_0 + \beta_1x_{1i}+ \cdots+\beta_kx_{ki}  + \log(t_i)\\
\end{aligned}
$$



--- .class #id


## The Offset

- $\log(t_i)$ is called an offset. 
- We fit a regression model and fix the offset coefficient so that it is 1.


--- .class #id

## Rationale for $\log()$ transform

- Many times we call Poisson regression, log-linear regression. The rationale behind using the log transform is:
    - $\log(\lambda)$ has a range of $-\infty$ to $\infty$ even though $\lambda>0$. This means there are no restrictions to a specific range. 
    - Maximum likelihood estimation works extremely well with the $\log()$ relationship. This is due to the fact that the $\log()$ link is something called the canonical link between outcome and covariates. 


--- .class #id



## Analysis of Grouped or Individual Data

We can actually enter data in  different ways

1. Individual Data: one line per patient. 
2. Grouped Data: grouping data by a covariate pattern. 
    + This happens when all covariates are categorical
    + There are no differences with inferences in either case. 


--- .class #id


## Interpreting Coefficients

- Given our data we consider the following model:
$$\log\left(\lambda_{x_1,x_2}\right)= \beta_0 + \beta_1x_1 + \beta_2x_2$$
- where
    - $X_1$ is Alcohol Use
        + 1 for Daily
        + 0 for Less than Daily
    - $X_2$ is mean-centered age at baseline



--- .class #id

## What is $\beta_0$?



- $\beta_0$ can be interpreted as:
$$\beta_0 = \log\left(\lambda_{x_1=0, x_2=0}\right)$$
  + This then represents the log CRC rate for less than daily drinkers who are at the mean age.
  + The CRC rate for less than daily drinkers who are at the mean age is $\exp(\beta_0)$. 


--- .class #id


##  What is $\beta_1$?

-  $\beta_1$ can be interpreted as:
  + Consider 2 subjects who are the same age but differ in drinking status:

$$\begin{aligned}
\log\left(\lambda_{x_1=0, x_2}\right) &= \beta_0 + \beta_2x_2\\
\log\left(\lambda_{x_1=1, x_2}\right) &= \beta_0 + \beta_1+  \beta_2x_2\\
\beta_1 &= \log\left(\lambda_{x_1=1, x_2}\right) - \log\left(\lambda_{x_1=0, x_2}\right)\\
&= \log\left( \dfrac{\lambda_{x_1=1, x_2}}{\lambda_{x_1=0, x_2}}\right)\\
\end{aligned}$$


--- .class #id

## How do you interpret $\beta_1$?

  + This is the log CRC rate ratio comparing daily drinking to less than daily drinking in subjects who are the same age. 
  + The CRC rate ratio comparing daily drinking to less than daily drinking in subjects who are the same age is $\exp(\beta_1)$.
  

--- .class #id

##   What is $\beta_2$?
    
3.  What is $\beta_2$?:
  + Consider 2 subjects who differ in age by one year but have the same drinking status:

$$\begin{aligned}
\log\left(\lambda_{x_1, x_2}\right) &= \beta_0 + \beta_1 x_1 +  \beta_2x_2\\
\log\left(\lambda_{x_1, x_2+1}\right) &= \beta_0 + \beta_1 x_1 +  \beta_2(x_2+1)\\
\beta_2 &= \log\left(\lambda_{x_1, x_2+1}\right) - \log\left(\lambda_{x_1, x_2}\right)\\
&= \log\left( \dfrac{\lambda_{x_1, x_2+1}}{\lambda_{x_1, x_2}}\right)\\
\end{aligned}$$


--- .class #id


## How do you interpret $\beta_2$?

  + This is the log CRC rate ratio comparing a one year increase over mean age for patients who have the same drinking status.  
  + The CRC rate ratio comparing a one year increase over mean age for subjects who have the same drinking status is $\exp(\beta_2)$.
  

--- .class #id


## Model in R


```{r}
phscrc$mean.cent.age <- phscrc$age - mean(phscrc$age, na.rm=TRUE)
fit5 <- glm(crc~alcohol.use + mean.cent.age + offset(log(cayrs)), data=phscrc, family=poisson(link='log'))
```
  
```{r,echo=FALSE}
library(broom)
tidy5 <- tidy(fit5, exponentiate=T, conf.int=T)[,-c(3:4)]
              
 
knitr::kable(tidy5)
```


--- .class #id


## Interpretation of Coefficients



- ***Intercept***:
    + The CRC rate for less than daily drinkers who are 53 years old is 0.001. 
- ***Alcohol Use***:
    + The CRC rate ratio comparing daily drinking to less than daily drinking in subjects who are the same age is 1.413 although it is insignificant.
    + The CRC rate for daily drinkers is 41.3% greater than the CRC rate of less than daily drinkers  although it is insignificant.
    


--- .class #id


## Interpretation of Coefficients


- ***Age***:
    + The CRC rate ratio comparing a one year increase over mean age for subjects who have the same drinking status is 1.08.
    + The CRC rate for one year increase in mean age is 8% larger than the CRC rate for subjects at the mean age and who have the same drinking status. 
    

--- .segue bg:grey


# Model fit for Poisson Regression


--- .class #id

## Deviance Goodness-of-Fit Test

- **Deviance** is a a measure of how close our model predicts the actual observed outcomes. 
    + *This is similar to residuals in linear regression*
- We can use this as a basic test for goodness of fit since we hope our predictions are close to actual outcomes.


--- .class #id

## Distribution

- We first must understand what the distribution of this would be
    + If our model is correctly specified we must determine how much variation we expect in the observed outcomes around the predicted means under the assumption that our data is Poisson. 
    + It can be shown that deviance follows a $\chi^2$ distribution equal to the difference in parameters between the model fit at the saturated model, $n-p$.
    


--- .class #id

## Chi-Square Test

- This means we can use a $\chi^2$ test for this with the hypothesis of:

$$H_0:\text{ The Model is Correctly Specified }$$
$$\text{vs.}$$
$$H_1:\text{ The Model is Not Correctly Specified }$$

```{r}
pchisq(fit5$deviance, df=fit5$df.residual, lower.tail=FALSE)
```



--- .class #id

## Overdispersion

-When we deal with Poisson data we are saying that
$$E(X) = Var(X)$$

- In other words we are saying that the mean is equal to the variance. If this is not true:
    - We still have valid estimates of relevant event rates
    - We tend to underestimate variance and then have p-values that are too small and confidence intervals that are too narrow. 
    - We correct for this using Negative-Binomial Regression. 


--- .class #id


## Dispersion Test

We can test for this in R:

```{r}
library(AER)
dispersiontest(fit5)
```


--- .class #id

## Results of Dispersion Test

- From this test we see that our dispersion is 1 and so we have correctly scaled the variance compared to the mean. 
- If it was not we could make a change that would correct it without having to learn a new regression model:



--- .class #id


##  Results

```{r}
summary(fit5)
```


--- .class #id

## Results Correcting for Dispersion

```{r}
summary(fit5, dispersion=0.9841428)
```

